---
title: "Classification with PLS or RF"
description: >
  Classify samples with a core method — the same recursive-elimination workflow as
  the regression article, but with a categorical target.
---

```{r, include = FALSE}
source("_common.R")
```

Classification in MUVR2 comes in two flavours. This article uses a **core method**
(PLS or random forest) through `MUVR2()`; the [elastic net](classification-en.html)
version, `MUVR2_EN()`, behaves differently enough to need its own walkthrough. With
a core method the analysis is the same repeated double cross-validation with
recursive variable elimination used for [regression](regression-pls.html) — the only
change is that the target is a categorical variable.

We use the `mosquito` data: 16S rRNA microbiota composition of 29 *Anopheles
gambiae* mosquitoes caught in three villages in western Burkina Faso. The question
is whether a mosquito's microbiota tells you which village it came from.

```{r setup}
library(MUVR2)
library(doParallel)

data("mosquito")
```

`mosquito` loads a categorical target `Yotu` (three villages) and a numeric matrix
`Xotu` of 1678 operational taxonomic units (OTUs).

```{r data}
dim(Xotu)
table(Yotu)
```

That class table matters for one setting: as a rule, **`nOuter` should not exceed
the size of the smallest class**, so that every class can appear in every segment.
Here the smallest class has 8 members, so `nOuter` ≤ 8.

## Fitting the model

The call is the regression call with a categorical `Y`. `DA = TRUE` asks for a
discriminant (i.e. classification) analysis; it is inferred anyway when `Y` is a
factor or character vector. Switch `method` to `"RF"` for a random forest core.

```{r fit}
classModel <- MUVR2(X =         Xotu,
                    Y =         Yotu,
                    nRep =      4,     # tutorial uses 35
                    nOuter =    6,     # tutorial uses 8
                    varRatio =  0.75,
                    method =    "PLS",
                    DA =        TRUE,
                    modReturn = TRUE)
```

## What came out

Classification uses different fitness metrics from regression — balanced error rate
(BER) by default, with the number of misclassifications available too. Unlike
elastic net, a core method **does** do recursive elimination, so you get the full
`min`/`mid`/`max` triple: one BER, one misclassification count and one variable
count for each.

```{r metrics}
classModel$ber    # balanced error rate for the min, mid and max models
classModel$miss   # misclassifications
classModel$nVar   # number of variables in each
```

```{r classes}
head(cbind.data.frame(actual = Yotu, predicted = classModel$yClass$mid))
```

## Validation: how many variables?

`ggplotVAL()` shows the validation metric — here BER — against the number of
variables surviving the recursive elimination, with the `min`, `mid` and `max`
selections marked.

```{r val}
ggplotVAL(classModel)
```

## Predictions: the swimlane plot

For classification, `ggplotMV()` gives a "swimlane" plot: each lane is one
observation, the class prediction probabilities are colour-coded and nudged apart
by class, small dots are individual repetitions and large dots the consensus, and
misclassified observations are ringed at the probability of their true class. With
a core method you can switch between the three models with the `model` argument.

```{r mv, fig.width = 9, fig.height = 5}
ggplotMV(classModel, model = "mid")
```

## Stability

Same as regression, except the bottom panels now show misclassifications and BER
instead of Q2. Watch the cumulative curves: once they flatten, `nRep` is high
enough.

```{r stability, fig.height = 8}
ggplotStability(classModel, model = "mid")
```

## Which variables?

Because a core method ranks variables by recursive elimination, `ggplotVIRank()`
draws the familiar rank boxplot — lower is a more reproducible predictor — rather
than the elastic net selection map.

```{r virank, fig.height = 6}
ggplotVIRank(classModel, model = "mid", n = 20)
```

```{r getvirank}
head(getVIRank(classModel, model = "mid"))
```

## Next steps

- Elastic net is the other way to classify, and the only core that supports
  [covariate adjustment](covariate-adjustment.html); see
  [classification with elastic net](classification-en.html).
- Before believing any of this, check the model against the null with
  [resampling tests](resampling-tests.html).
- When individuals contribute several samples, see
  [repeated samples](multilevel.html).
