---
title: "Classification with elastic net"
description: >
  Classify samples with the elastic net core, and pick variables from a model
  that does not do recursive elimination.
---

```{r, include = FALSE}
source("_common.R")
```

This article walks through a classification analysis with an elastic net (EN)
core, using the `mosquito` data: 16S rRNA microbiota composition of 29
*Anopheles gambiae* mosquitoes caught in three villages in western Burkina Faso.
The question is whether a mosquito's microbiota tells you which village it came
from.

Elastic net behaves differently enough from PLS and RF that it needs its own
walkthrough, even though the outline of the analysis is the same.

```{r setup}
library(MUVR2)
library(doParallel)

data("mosquito")
```

`mosquito` loads a categorical target `Yotu` (three villages) and a numeric matrix
`Xotu` of 1678 operational taxonomic units (OTUs).

```{r data}
dim(Xotu)
table(Yotu)
```

That class table is worth a moment: as a rule, **`nOuter` should not exceed the
size of the smallest class**, so that every class can appear in every segment.
Here the smallest class has 8 members, so `nOuter` ≤ 8.

## Fitting the model

Use `MUVR2_EN()` rather than `MUVR2()`. There is no `method` argument — the
function only does elastic net. `DA = TRUE` asks for a discriminant (i.e.
classification) analysis, though it is inferred anyway when `Y` is a factor or
character vector.

```{r fit}
classModel <- MUVR2_EN(X = Xotu,
                       Y = Yotu,
                       nRep = nRep_article,      # tutorial uses nRep = 35
                       nOuter = nOuter_article,  # tutorial uses nOuter = 8
                       DA = TRUE,
                       modReturn = TRUE)
```

## Why elastic net needs `getVar()`

PLS and RF whittle variables away by recursive elimination, which produces a
continuous validation curve and hence an obvious place to cut. **Elastic net does
no such elimination.** It ranks a variable by how often it received a non-zero
beta coefficient across the `nRep` × `nOuter` calibration models — and that is all
it gives you. There is no curve, so `MUVR2_EN()` deliberately does not report a
variable selection:

```{r no-nvar}
is.null(classModel$Var)
head(classModel$varTable)  # times each variable had a non-zero coefficient
```

Variable selection is a separate, deliberate step, done with `getVar()`, and it
offers two options.

### Option 1: `option = "fitness"`

Smooth a curve through the `nRep` × `nOuter` fitness estimates, then cut it the
same way PLS and RF are cut — at the edges of the region within `robust` (5% by
default) of the best value.

```{r getvar-fitness}
classModel <- getVar(classModel, option = "fitness", robust = 0.05, outlier = "none")
classModel$nVar
```

**Look at the curve before you trust it.** Each point is one calibration model, so
coverage across the range of variable counts is uneven, and with too few
repetitions the smoothed curve will not show the U-shape it should:

```{r val-fitness, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVAL(classModel)
ggplotVAL(classModel)
```

If the curve is not U-shaped, that is the model telling you `nRep` (and possibly
`nOuter`) is too low. This article's model is small, so do not be surprised if the
curve here is unconvincing — with `nRep = 35, nOuter = 8` and its 280 underlying
fitness estimates, the tutorial's version has a well-defined global minimum.

Other levers: raise or lower `robust` to change the slack allowance, set
`outlier = "IQR"` or `outlier = "residual"` to drop outlying points that make the
curve squiggle, and lower `span` to reduce smoothing.

### Option 2: `option = "quantile"`

If no amount of tuning produces a usable curve, ignore performance altogether and
select directly from the *distribution* of how many variables the calibration
models chose. `mid` is the median; `quantile` sets where `min` and `max` fall.

```{r getvar-quantile, fig.show = "hold", out.width = "49%", fig.width = 5.5}
classModelQ <- getVar(classModel, option = "quantile", quantile = 0.25, outlier = "none")
classModelQ$nVar

plotVAL(classModelQ)
ggplotVAL(classModelQ)
```

This puts the selections where the data actually is, at the cost of no longer
taking prediction performance into account. It is a fallback, not a default.

## What came out

Classification uses different fitness metrics from regression — balanced error
rate (BER) by default. And note the other elastic net difference: because variable
selection happens *after* fitting, there is only **one** model, hence one BER and
one set of predictions, not one each for min/mid/max.

```{r metrics}
classModel$ber    # one balanced error rate, not three
classModel$miss   # number of misclassifications
classModel$auc    # AUROC per class
classModel$nVar   # variables in the min, mid and max selections
```

```{r classes}
head(cbind.data.frame(actual = Yotu, predicted = classModel$yClass))
```

## Predictions: the swimlane plot

For classification, `plotMV()` gives a "swimlane" plot. Each lane is one
observation. Within a lane, the class prediction probabilities are colour-coded and
nudged apart by class; small dots are individual repetitions, large dots the
consensus. Misclassified observations are ringed at the probability of their
*true* class.

Reading it: a confident, correct prediction is one where the true class's dot sits
high and alone, with its repetitions tightly clustered. A wide spread of small dots
means the prediction moved a lot between repetitions.

```{r mv, fig.width = 9, fig.height = 5}
plotMV(classModel)
```

```{r mv-gg, fig.width = 9, fig.height = 5}
ggplotMV(classModel)
```

## Stability

Same as regression, except the bottom panels now show misclassifications and BER
instead of Q2.

```{r stability, fig.height = 8}
ggplotStability(classModel, model = "mid")
```

## Which variables?

`plotVIRank()` is also different for elastic net, and necessarily so: there are no
ranks per repetition to draw a boxplot of. Instead each row is one calibration
model, each column a variable, and red means that variable had a non-zero
coefficient in that model. Models are ordered from most to fewest variables
selected, variables by selection rate.

What you are looking for is a solid red block on the left: a core set of variables
that got picked regardless of how many variables the model felt like selecting.

```{r virank, fig.width = 8, fig.height = 5}
plotVIRank(classModel, model = "mid", maptype = "heatmap")
```

```{r virank-gg, fig.width = 8, fig.height = 5}
ggplotVIRank(classModel, model = "mid", maptype = "heatmap")
```

```{r getvirank}
head(getVIRank(classModel, model = "mid"))
```

## Next steps

- Elastic net is the only core that supports
  [covariate adjustment](covariate-adjustment.html) — that is the main reason to
  reach for it.
- Check the model against the null with
  [resampling tests](resampling-tests.html).
