---
title: "Covariate adjustment"
description: >
  Force a covariate into every model, so that what survives is what predicts the
  target over and above it.
---

```{r, include = FALSE}
source("_common.R")
```

Suppose age predicts your outcome, and half of your metabolites correlate with age.
A plain variable selection will happily hand you those metabolites, and you will
have discovered age.

MUVR2 lets you adjust for a covariate by **suppressing regularization on it**: the
covariate is exempted from the elastic net penalty, so it is in every model,
always, for free. Every other variable then has to earn its place by predicting the
target *over and above* what the covariate already explains. Variables that were
only riding on the covariate lose their rank and stop being selected.

This is available in **`MUVR2_EN()` only**. There is no equivalent for the PLS or
random forest cores.

Doing it is almost anticlimactic — pass `keep`:

```r
MUVR2_EN(X = X, Y = Y, keep = c("age", "sex"), ...)
```

The names in `keep` must match column names in `X` exactly.

## A worked example

We will reuse the mosquito classification from the
[elastic net article](classification-en.html). There is no real covariate in that
data, so we will manufacture one: pick a predictor, declare it the covariate, and
watch what happens to the variables correlated with it. (Not a real research
question — but it makes the mechanism visible.)

```{r setup}
library(MUVR2)
data("mosquito")

classModel <- MUVR2_EN(X = Xotu,
                       Y = Yotu,
                       nRep = nRep_article,
                       nOuter = nOuter_article,
                       DA = TRUE,
                       modReturn = TRUE)
classModel <- getVar(classModel)
```

### Choosing the covariate

A covariate should not be a *strong* predictor of the target. If it is, the model is
pushed toward a Hauck–Donner effect and the algorithm will in practice throw an
error. So: pick something inside the `max` selection but not near the top of the
ranking.

```{r rank-before}
ranks_before <- getVIRank(classModel, model = "max")
head(ranks_before)

covariate <- "OTU_100"
ranks_before[covariate, ]
```

Now find what correlates with it, among the selected variables:

```{r correlated}
max_predictors <- Xotu[, classModel$Var$max]
cors <- cor(max_predictors)[, covariate]
correlated <- cors[abs(cors) > 0.75 & names(cors) != covariate]
correlated

## and how strong that correlation is
if (length(correlated) > 0) {
  partner <- names(correlated)[1]
  cor.test(Xotu[, covariate], Xotu[, partner])$p.value
}
```

The tutorial's full-scale model (`nRep = 35, nOuter = 8`) finds `OTU_1962`
correlating with `OTU_100` at r = 0.83, p < 0.001. Whether the smaller model on this
page reproduces exactly that depends on which variables land in its `max` selection.

### Refitting with the covariate held in

```{r fit-adjusted}
classModel2 <- MUVR2_EN(X = Xotu,
                        Y = Yotu,
                        keep = c(covariate),   # <- the adjustment
                        nRep = nRep_article,
                        nOuter = nOuter_article,
                        DA = TRUE,
                        modReturn = TRUE)
classModel2 <- getVar(classModel2)
```

### What changed

Two things move for a variable that was only predicting the target *through* the
covariate: its **rank drops**, and the **number of calibration models that select
it falls**, often to zero.

```{r compare}
ranks_after <- getVIRank(classModel2, model = "max")

compareVar <- function(v) {
  data.frame(
    variable = v,
    rank_before = if (v %in% rownames(ranks_before)) ranks_before[v, "rank"] else NA,
    rank_after = if (v %in% rownames(ranks_after)) ranks_after[v, "rank"] else NA,
    selected_before = ifelse(is.na(classModel$varTable[v]), 0, classModel$varTable[v]),
    selected_after = ifelse(is.na(classModel2$varTable[v]), 0, classModel2$varTable[v]),
    row.names = NULL
  )
}

do.call(rbind, lapply(names(correlated), compareVar))
```

In the tutorial's full-scale model, `OTU_1962` falls from rank 20 to rank 703, and
from being selected in 150 of the 280 calibration models to being selected in none
of them. The numbers here will be smaller and noisier, because this model is
smaller — but the direction is the story.

Rank and selection counts fluctuate between runs anyway, thanks to the random
segment sampling. Do not over-read a small movement in a single variable.

## What to take away

- `keep` exempts a variable from regularization; it does not remove it. It is always
  in the model, and everything else is judged relative to it.
- Use it when you have a genuine confounder you can name — age, sex, batch, BMI —
  and you want associations that survive it.
- It is elastic net only.
- The covariate must not be a strong predictor of the target.
