---
title: "Repeated samples and multilevel analysis"
description: >
  Two different kinds of sample dependency, and the two different things you have
  to do about them.
---

```{r, include = FALSE}
source("_common.R")
```

Both of the other analysis articles assume one sample per individual. When that is
not true, cross-validation will happily put two samples from the same person in
both the training and the test set, and the model will look better than it is.

There are two distinct situations here, and they need different fixes.

## 1. Repeated samples: use `ID`

If the same individual contributed several samples but there is no systematic
effect distinguishing them — e.g. two measurement occasions, both under
free-living conditions — the dependency is a nuisance, not a signal. The fix is to
keep an individual's samples together when the data is split into segments, which
is what the `ID` argument does.

The `freelive` data is `freelive2` with the repeats put back: 112 samples from 58
individuals, with `IDR` saying which sample came from whom.

```{r setup}
library(MUVR2)

data("freelive")

dim(XRVIP)         # 112 samples this time, not 58
length(unique(IDR))
```

The only change to the model call is `ID = IDR`:

```{r fit-id}
regrModel <- MUVR2(X = XRVIP,
                   Y = YR,
                   ID = IDR,                 # <- keeps an individual's samples together
                   nRep = nRep_article,
                   nOuter = nOuter_article,
                   varRatio = varRatio_article,
                   method = "PLS",
                   modReturn = TRUE)

regrModel$fitMetric
```

`ID` works with every core method (`MUVR2()` with PLS or RF, and `MUVR2_EN()`) and
with every problem type. If your samples are not independent, use it.

## 2. A systematic within-individual effect: multilevel analysis

Now suppose the dependency *is* the signal: a crossover trial where each subject
received both treatments, or a before/after intervention. Here `ID` alone is not
enough, because the between-individual variation — which is usually large, and
which you do not care about — gets conflated with the within-individual treatment
effect, which is the whole question.

If a standard classification is the multivariate analogue of an unpaired t-test,
multilevel analysis is the analogue of a **paired** one.

Mechanically, multilevel analysis is not a new algorithm at all. It is a
preprocessing trick: instead of modelling the two conditions as separate
observations, you model their *difference*.

$$EM = X_B - X_A \qquad \text{(or } EM = \log(X_B / X_A) \text{)}$$

That **effect matrix** becomes the predictor matrix, and the target is a dummy
variable that MUVR2 constructs internally. **You must compute the effect matrix
yourself**; MUVR2 will not do it for you. Then pass it as `X` with `ML = TRUE`,
and pass no `Y` at all.

(This is unrelated to "multilevel models" in the classical-statistics sense of
random effects. Same word, different thing.)

The `crisp` data is a crossover study: untargeted plasma metabolomics from 21
subjects who each received two different breakfast interventions. `crispEM` is
already an effect matrix — 21 rows, one per individual, 1508 columns of
differences in area-under-the-curve between the two meals.

```{r ml-data}
data("crisp")
dim(crispEM)
```

```{r ml-fit}
MLModel <- MUVR2(X = crispEM,
                 ML = TRUE,      # <- multilevel; note there is no Y argument
                 nRep = nRep_article,
                 nOuter = nOuter_article,
                 varRatio = varRatio_article,
                 method = "RF",
                 modReturn = TRUE)
```

```{r ml-metrics}
MLModel$nVar       # variables in the min, mid and max models
MLModel$miss       # misclassified observations
MLModel$ber        # balanced error rate
MLModel$fitMetric  # R2/Q2 of the underlying dummy regression
```

## Reading the multilevel prediction plot

This one deserves care, because it looks like neither the regression plot nor the
classification swimlane.

Internally, MUVR2 models both the effect matrix and its negation, so each of the
21 individuals appears **twice**: once in the top half with an expected dummy value
of −1, once in the bottom half with +1. The decision boundary is at 0.

So: points in the top half should fall left of zero, points in the bottom half
right of zero. Any point on the wrong side of the dashed vertical line is a
misclassification. And this is why `MLModel$miss` reports twice as many misses as
there are misclassified individuals — it counts both halves.

```{r ml-mv, fig.height = 6}
ggplotMV(MLModel, model = "min")
```

## Everything else behaves as usual

The validation plot is read exactly as in the [regression article](regression-pls.html):

```{r ml-val}
ggplotVAL(MLModel)
```

The stability plot gives Q2 (of the dummy regression), misclassifications and BER:

```{r ml-stability, fig.height = 8}
ggplotStability(MLModel, model = "min")
```

And variable ranks come out as the familiar boxplot:

```{r ml-virank, fig.height = 6}
ggplotVIRank(MLModel, model = "min")
```

```{r ml-getvirank}
head(getVIRank(MLModel, model = "min"))
```

## Recap

Any of the three cores — PLS, RF, elastic net — can do any of the three problem
types: regression, classification, multilevel. And all of them accept `ID`.

What differs between the cores is *how variables get selected*: `MUVR2()` (PLS and
RF) does it by recursive elimination during fitting, and therefore reports separate
predictions and fitness for min, mid and max. `MUVR2_EN()` does it afterwards, from
how often each variable earned a non-zero coefficient, and therefore reports one
prediction and one fitness — with min/mid/max variable selections layered on top by
[`getVar()`](classification-en.html).
