---
title: "Plot gallery: base vs ggplot2 vs plotly"
description: >
  Every MUVR2 plot, in all three flavours, side by side.
---

```{r, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(fig.height = 4.5)
```

Every plot in MUVR2 comes in two flavours:

- **`plotMV()`, `plotVAL()`, …** — base graphics. Draws to the device, returns
  nothing. This is what the package has always had.
- **`ggplotMV()`, `ggplotVAL()`, …** — the same plot, returning a `ggplot` object.
  Theme it, compose it, `ggsave()` it — and because it is a real object, hand it to
  `plotly::ggplotly()` and it becomes interactive for free.

They draw the same information. Where they differ deliberately, this page says so.

```{r setup}
library(MUVR2)
library(ggplot2)
library(plotly)

data("freelive2")
data("mosquito")
data("crisp")
```

```{r models}
regrModel <- MUVR2(X = XRVIP2, Y = YR2,
                   nRep = nRep_article, nOuter = nOuter_article,
                   varRatio = varRatio_article, method = "PLS", modReturn = TRUE)

classModel <- MUVR2_EN(X = Xotu, Y = Yotu,
                       nRep = nRep_article, nOuter = nOuter_article,
                       DA = TRUE, modReturn = TRUE)
classModel <- getVar(classModel)

MLModel <- MUVR2(X = crispEM, ML = TRUE,
                 nRep = nRep_article, nOuter = nOuter_article,
                 varRatio = varRatio_article, method = "RF", modReturn = TRUE)
```

## plotMV — predictions

### Regression

Actual target on x, prediction on y. Grey dots are individual repetitions, black
dots the consensus.

```{r mv-regr, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotMV(regrModel, model = "min")
ggplotMV(regrModel, model = "min")
```

Hover to read off individual samples:

```{r mv-regr-plotly}
ggplotly(ggplotMV(regrModel, model = "min"))
```

### Classification — the swimlane plot

One lane per observation; class probabilities colour-coded and nudged apart.
Misclassified samples are ringed at the probability of their true class.

This is the plot where interactivity earns its keep: with 29 lanes the base plot is
already crowded, and hovering tells you which sample and which class each dot
belongs to.

```{r mv-class, fig.width = 9, fig.height = 5}
plotMV(classModel)
```

```{r mv-class-gg, fig.width = 9, fig.height = 5}
ggplotMV(classModel)
```

```{r mv-class-plotly, fig.width = 9}
ggplotly(ggplotMV(classModel))
```

### Multilevel

Each individual appears twice — once per half of the effect matrix — with expected
values of −1 (top) and +1 (bottom), and a decision boundary at 0.

```{r mv-ml, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 6}
plotMV(MLModel, model = "min")
ggplotMV(MLModel, model = "min")
```

## plotVAL — validation metric vs number of variables

### PLS / RF

Light grey: individual inner segments. Dark grey: averaged per repetition. Black:
averaged over everything. Vertical lines: the min, mid and max selections.

```{r val-pls, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVAL(regrModel)
ggplotVAL(regrModel)
```

```{r val-pls-plotly}
ggplotly(ggplotVAL(regrModel))
```

### Elastic net

Different plot, different meaning: each point is one calibration model, and the
curve is the fitness curve `getVar()` smoothed through them.

```{r val-en, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVAL(classModel)
ggplotVAL(classModel)
```

## plotVIRank — variable importance

### PLS / RF: rank boxplots

Each box is a variable's rank across repetitions; lower is better. A tight box near
the left is a reproducible predictor. Yellow (base) / highlighted (ggplot2) marks
the variables inside the selection.

```{r virank-pls, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 6}
plotVIRank(regrModel, n = 20)
ggplotVIRank(regrModel, n = 20)
```

```{r virank-pls-plotly, fig.height = 6}
ggplotly(ggplotVIRank(regrModel, n = 20))
```

### Elastic net: a selection map

Rows are calibration models, columns are variables, red means selected. Long
variable names are exactly the case where hovering beats squinting at an axis.

```{r virank-en, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVIRank(classModel, model = "mid", maptype = "heatmap")
ggplotVIRank(classModel, model = "mid", maptype = "heatmap")
```

```{r virank-en-dot, fig.width = 8}
ggplotly(ggplotVIRank(classModel, model = "mid", maptype = "dotplot"))
```

## plotStability — did it converge?

The base version stacks its panels with `par(mfrow=)`; the ggplot2 version facets
them, which means the panels share an x-axis and stay aligned when you resize.

Watch the *cumulative* (black) series: once it is flat, `nRep` is high enough.

```{r stability, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 7}
plotStability(regrModel, model = "min")
ggplotStability(regrModel, model = "min")
```

```{r stability-plotly, fig.height = 7}
ggplotly(ggplotStability(regrModel, model = "min"))
```

## plotPerm — model fitness vs the null

One deliberate difference here. The base version draws the Student's t curve on a
*second, hidden axis*, so the curve and the histogram are not on a common scale.
The ggplot2 version puts both on the density scale, so they can be read against
each other.

```{r perm, fig.show = "hold", out.width = "49%", fig.width = 5.5}
set.seed(11)
distribution <- rnorm(200, mean = 0.15, sd = 0.08)
actual <- 0.42

plotPerm(actual, distribution, type = "t", xlab = "Q2")
ggplotPerm(actual, distribution, type = "t", xlab = "Q2")
```

Several tests at once:

```{r perm-multi, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPerm(actual, distribution, multiple_p_shown = c("t", "smooth"), xlab = "Q2")
ggplotPerm(actual, distribution, multiple_p_shown = c("t", "smooth"), xlab = "Q2")
```

See [resampling tests](resampling-tests.html) for what these p-values mean and
which one to trust.

## plotPCA

`plotPCA()` takes a `file` argument and writes a png as a side effect.
`ggplotPCA()` does not: it returns the object, and you save it with `ggsave()`.

```{r pca, fig.show = "hold", out.width = "49%", fig.width = 5.5}
pca <- prcomp(XRVIP2)

plotPCA(pca, colVar = YR2)
ggplotPCA(pca, colVar = YR2)
```

```{r pca-plotly}
ggplotly(ggplotPCA(pca, colVar = YR2))
```

## biplotPLS

Scores and loadings in one panel: scores on the primary axes, loadings on the
secondary ones.

A caveat worth knowing: `plotly` does not render ggplot2's secondary axes, so the
interactive version loses the loadings axis labels even though the arrows are still
correctly placed and still hoverable. Use the static version when the loadings scale
matters.

```{r biplot, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 5.5}
PLSfit <- regrModel$Fit$plsFitMin

biplotPLS(PLSfit, comps = 1:2, xCol = YR2, labPlSc = FALSE, labPlLo = FALSE)
ggbiplotPLS(PLSfit, comps = 1:2, xCol = YR2, labPlSc = FALSE, labPlLo = FALSE)
```

## plotPred

```{r pred, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPred(YR2, regrModel$yPred)
ggplotPred(YR2, regrModel$yPred)
```

## Customising the ggplot2 versions

The point of returning an object is that it is yours to change afterwards:

```{r customise}
ggplotMV(regrModel, model = "min") +
  labs(title = "Wholegrain rye intake: predicted vs actual",
       subtitle = "Grey: individual repetitions. Black: consensus.",
       caption = "MUVR2-PLS, minimal-optimal model") +
  theme_minimal(base_size = 13)
```

And saving one:

```r
p <- ggplotVIRank(regrModel, n = 20)
ggsave("virank.png", p, width = 8, height = 6, dpi = 300)
```
