---
title: "Plot gallery: base vs ggplot2 vs plotly"
description: >
  Every MUVR2 plot, in all three flavours, side by side.
---

```{r, include = FALSE}
source("_common.R")
knitr::opts_chunk$set(fig.height = 4.5)
```

Every plot in MUVR2 comes in two flavours:

- **`plotMV()`, `plotVAL()`, …** — base graphics. Draws to the device, returns
  nothing. This is what the package has always had.
- **`ggplotMV()`, `ggplotVAL()`, …** — the same plot, returning a `ggplot` object.
  Theme it, compose it, `ggsave()` it — and because it is a real object, hand it to
  `plotly::ggplotly()` and it becomes interactive for free.

They draw the same information. Where they differ deliberately, this page says so.

```{r setup}
library(MUVR2)
library(ggplot2)
library(plotly)

data("freelive2")
data("mosquito")
data("crisp")
```

```{r models}
regrModel <- MUVR2(X =         XRVIP2,
                   Y =         YR2,
                   nRep =      4,     # tutorial uses 35
                   nOuter =    6,     # tutorial uses 8
                   varRatio =  0.75,
                   method =    "PLS",
                   modReturn = TRUE)

classModel <- MUVR2_EN(X =         Xotu,
                       Y =         Yotu,
                       nRep =      4,     # tutorial uses 35
                       nOuter =    6,     # tutorial uses 8
                       DA =        TRUE,
                       modReturn = TRUE)
classModel <- getVar(classModel)

MLModel <- MUVR2(X =         crispEM,
                 ML =        TRUE,
                 nRep =      4,     # tutorial uses 35
                 nOuter =    6,     # tutorial uses 8
                 varRatio =  0.75,
                 method =    "RF",
                 modReturn = TRUE)
```

## plotMV — predictions

### Regression

Actual target on x, prediction on y. Grey dots are individual repetitions, black
dots the consensus.

```{r mv-regr, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotMV(regrModel, model = "min")
ggplotMV(regrModel, model = "min")
```

Hover to read off individual samples — pass `tooltip = "text"` so the tooltip
shows the sample name rather than the raw coordinates:

```{r mv-regr-plotly}
ggplotly(ggplotMV(regrModel, model = "min"), tooltip = "text")
```

### Classification — the swimlane plot

This plot confuses people, so it is worth being precise about what a dot is.

The x-axis is the sample (one lane per observation). For a *3-class* problem, the
model does not output a single prediction per sample — it outputs, for each sample,
a **probability of belonging to each of the three classes** (which sum to about 1).
So every sample gets three coloured dots, one per class, and they are nudged apart
horizontally within the lane so they do not sit on top of each other. A confident,
correct prediction is a lane where the true class's dot sits high and alone.

On top of that, each of those three positions carries both the **consensus**
prediction (the larger dot) and the prediction from **each repetition** (the
smaller dots) — so the spread of small dots around a large one shows how much that
probability moved between repetitions. Misclassified samples are ringed at the
probability of their true class.

That is why a single lane can hold many dots: (3 classes) × (1 consensus + several
per-repetition), colour-coded by class.

This is the plot where interactivity earns its keep: with 29 lanes the base plot is
already crowded, and hovering tells you which sample and which class each dot
belongs to.

```{r mv-class, fig.width = 9, fig.height = 5}
plotMV(classModel)
```

```{r mv-class-gg, fig.width = 9, fig.height = 5}
ggplotMV(classModel)
```

```{r mv-class-plotly, fig.width = 9}
ggplotly(ggplotMV(classModel), tooltip = "text")
```

When the per-repetition cloud is more than you need, `consensusOnly = TRUE` drops
it and keeps only the consensus dots (and the misclassification rings). It works
for every model type, but the swimlane is where it helps most:

```{r mv-class-consensus, fig.width = 9, fig.height = 5}
ggplotMV(classModel, consensusOnly = TRUE)
```

### Multilevel

Each individual appears twice — once per half of the effect matrix — with expected
values of −1 (top) and +1 (bottom), and a decision boundary at 0.

```{r mv-ml, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 6}
plotMV(MLModel, model = "min")
ggplotMV(MLModel, model = "min")
```

```{r mv-ml-plotly, fig.height = 6}
ggplotly(ggplotMV(MLModel, model = "min"), tooltip = "text")
```

## plotVAL — validation metric vs number of variables

### PLS / RF

Light grey: individual inner segments. Dark grey: averaged per repetition. Black:
averaged over everything. Vertical lines: the min, mid and max selections.

```{r val-pls, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVAL(regrModel)
ggplotVAL(regrModel)
```

```{r val-pls-plotly}
ggplotly(ggplotVAL(regrModel))
```

### Elastic net

Different plot, different meaning: each point is one calibration model, and the
curve is the fitness curve `getVar()` smoothed through them.

```{r val-en, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVAL(classModel)
ggplotVAL(classModel)
```

```{r val-en-plotly}
ggplotly(ggplotVAL(classModel))
```

Selecting variables by quantile instead gives a third form of the plot — a
histogram of how many variables the calibration models chose:

```{r val-en-quantile, fig.show = "hold", out.width = "49%", fig.width = 5.5}
classModelQ <- getVar(classModel, option = "quantile")

plotVAL(classModelQ)
ggplotVAL(classModelQ)
```

```{r val-en-quantile-plotly}
ggplotly(ggplotVAL(classModelQ))
```

## plotVIRank — variable importance

### PLS / RF: rank boxplots

Each box is a variable's rank across repetitions; lower is better. A tight box near
the left is a reproducible predictor. These are drawn full width, not side by side,
because the variable names need the room to stay legible (`cut` trims them further).

With `n` at or below the model's variable selection, every variable shown is inside
the selection, so there is nothing to distinguish and the boxes are plain:

```{r virank-pls-base, fig.width = 9, fig.height = 6}
plotVIRank(regrModel, n = 20, cut = 12)
```

```{r virank-pls-gg, fig.width = 9, fig.height = 6}
ggplotVIRank(regrModel, n = 20, cut = 12)
```

The interactive version renders a real boxplot (an earlier version came out empty
through plotly), and hovering reads off the exact rank:

```{r virank-pls-plotly, fig.width = 9, fig.height = 6}
ggplotly(ggplotVIRank(regrModel, n = 20, cut = 12))
```

Ask for more variables than the model selected, and the ones inside the selection
turn yellow, the ones outside grey, with a line between them:

```{r virank-pls-wide-base, fig.width = 9, fig.height = 8}
n <- regrModel$nVar["min"] + 20
plotVIRank(regrModel, n = n, cut = 12)
```

```{r virank-pls-wide-gg, fig.width = 9, fig.height = 8}
ggplotVIRank(regrModel, n = n, cut = 12)
```

### Elastic net: a selection map

Rows are calibration models, columns are variables, red means selected. Long
variable names are exactly the case where hovering beats squinting at an axis.

Both `maptype`s exist in all three flavours. The heatmap:

```{r virank-en, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVIRank(classModel, model = "mid", maptype = "heatmap")
ggplotVIRank(classModel, model = "mid", maptype = "heatmap")
```

```{r virank-en-heat-plotly, fig.width = 8}
ggplotly(ggplotVIRank(classModel, model = "mid", maptype = "heatmap"))
```

And the dot plot, which unlike the heatmap leaves the variables in
selection-ratio order rather than clustering them. Note one deliberate
difference: `ggplotVIRank()` draws the *not selected* variables as faint grey
dots, where `plotVIRank()` draws them in white — that is, invisibly. Showing them
makes the grid readable as a grid.

```{r virank-en-dot, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotVIRank(classModel, model = "mid", maptype = "dotplot")
ggplotVIRank(classModel, model = "mid", maptype = "dotplot")
```

```{r virank-en-dot-plotly, fig.width = 8}
ggplotly(ggplotVIRank(classModel, model = "mid", maptype = "dotplot"))
```

## plotStability — did it converge?

The base version stacks its panels with `par(mfrow=)`; the ggplot2 version facets
them, which means the panels share an x-axis and stay aligned when you resize.
Both anchor each panel's y-axis the same way — a proportion runs 0 to 1 whether or
not the data goes near either end — so that a converged model does not look
unstable just because the axis zoomed into the noise.

Watch the *cumulative* (dark) series in each panel: once it is flat, `nRep` is high
enough.

```{r stability, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 7}
plotStability(regrModel, model = "min")
ggplotStability(regrModel, model = "min")
```

```{r stability-plotly, fig.height = 7}
ggplotly(ggplotStability(regrModel, model = "min"))
```

## plotPerm — model fitness vs the null

One deliberate difference here. The base version draws the Student's t curve on a
*second, hidden axis*, so the curve and the histogram are not on a common scale.
The ggplot2 version puts both on the density scale, so they can be read against
each other.

```{r perm, fig.show = "hold", out.width = "49%", fig.width = 5.5}
set.seed(11)
distribution <- rnorm(200, mean = 0.15, sd = 0.08)
actual <- 0.42

plotPerm(actual,
         distribution,
         type = "t",
         xlab = "Q2")
ggplotPerm(actual,
           distribution,
           type = "t",
           xlab = "Q2")
```

Several tests at once:

```{r perm-multi, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPerm(actual,
         distribution,
         multiple_p_shown = c("t", "smooth"),
         xlab =             "Q2")
ggplotPerm(actual,
           distribution,
           multiple_p_shown = c("t", "smooth"),
           xlab =             "Q2")
```

The interactive version keeps the p-value label (plotly drops plot annotations, so
`ggplotPerm()` draws it as a text layer instead) and lets you hover the histogram:

```{r perm-plotly, fig.width = 8}
ggplotly(ggplotPerm(actual,
                    distribution,
                    type = "t",
                    xlab = "Q2"),
         tooltip = "text")
```

See [resampling tests](resampling-tests.html) for what these p-values mean and
which one to trust.

## plotPCA

`plotPCA()` takes a `file` argument and writes a png as a side effect.
`ggplotPCA()` does not: it returns the object, and you save it with `ggsave()`.

The other difference is the colour legend. `plotPCA()` shows three swatches
labelled low / mid / high; `ggplotPCA()` shows a colour bar reading in the
variable's own units, so you can see not just which end is which but what value a
given sample has. `colVar` really is continuous — the base version cuts it into 40
shades and then labels only three of them.

`colLab` names the legend. Without it the legend is titled with whatever
expression you passed (here it would read `YR2`, which tells a reader nothing), so
it is worth setting to something meaningful:

```{r pca, fig.show = "hold", out.width = "49%", fig.width = 5.5}
pca <- prcomp(XRVIP2)

plotPCA(pca, colVar = YR2)
ggplotPCA(pca, colVar = YR2, colLab = "Rye intake")
```

```{r pca-plotly}
ggplotly(ggplotPCA(pca, colVar = YR2, colLab = "Rye intake"), tooltip = "text")
```

## biplotPLS

Scores and loadings in one panel: scores on the primary axes, loadings on the
secondary ones. The points are shaded by a continuous variable — here rye intake —
and `colLab` titles that legend, which the base version has no way to do.

A caveat worth knowing: `plotly` does not render ggplot2's secondary axes, so the
interactive version loses the loadings axis labels even though the arrows are still
correctly placed and still hoverable. Use the static version when the loadings scale
matters.

```{r biplot, fig.show = "hold", out.width = "49%", fig.width = 5.5, fig.height = 5.5}
PLSfit <- regrModel$Fit$plsFitMin

biplotPLS(PLSfit,
          comps =   1:2,
          xCol =    YR2,
          labPlSc = FALSE,
          labPlLo = FALSE)
ggbiplotPLS(PLSfit,
            comps =   1:2,
            xCol =    YR2,
            colLab =  "Rye intake",
            labPlSc = FALSE,
            labPlLo = FALSE)
```

## plotPred

```{r pred, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPred(YR2, regrModel$yPred)
ggplotPred(YR2, regrModel$yPred)
```

```{r pred-plotly}
ggplotly(ggplotPred(YR2, regrModel$yPred))
```

## Customising the ggplot2 versions

The point of returning an object is that it is yours to change afterwards:

```{r customise}
ggplotMV(regrModel, model = "min") +
  labs(title = "Wholegrain rye intake: predicted vs actual",
       subtitle = "Grey: individual repetitions. Black: consensus.",
       caption = "MUVR2-PLS, minimal-optimal model") +
  theme_minimal(base_size = 13)
```

And saving one:

```r
p <- ggplotVIRank(regrModel, n = 20)
ggsave("virank.png",
       p,
       width =  8,
       height = 6,
       dpi =    300)
```
