---
title: "Resampling and permutation tests"
description: >
  Is the model actually better than nothing? Build a null distribution, get a
  p-value, and check whether the modelling strategy is overfitting in the first
  place.
---

```{r, include = FALSE}
source("_common.R")
```

A Q2 of 0.4 or a balanced error rate of 0.3 means nothing on its own. The question
is what those numbers would have looked like if there were no association at all
between predictors and target. That is what this article is about, and you should
run it before believing any model.

The idea is old: scramble the target variable, refit, and see how well the model
does on nonsense. Repeat many times, and you have a null-hypothesis distribution
to compare the real model against.

## Permutation, and why MUVR2 resamples instead

A classic **permutation** test draws the null target by sampling the real target
*without replacement* — every null target is an exact reordering of the real one,
with exactly the same values in exactly the same proportions.

MUVR2 offers something slightly different by default, a **resampling** test. The
null target is drawn *from the empirical distribution* of the real target. In
regression that means values that were never observed but come from the same
distribution; in classification it means sampling classes according to their
observed probabilities (i.e. sampling with replacement).

The argument for it: a model's predictions are not confined to the exact values or
class proportions of the target it was trained on, so there is no reason the null
target should be either. Resampling puts a little more variability into the null,
which makes the resulting null distribution slightly broader and the p-values
slightly more conservative.

Both are available; `type` picks between them.

```{r setup}
library(MUVR2)
data("mosquito")
```

## The actual model

```{r actual}
model <- 1   # 1 = min, 2 = mid, 3 = max

MUVR2_actual <- MUVR2(X = Xotu,
                      Y = Yotu,
                      nRep = nRep_article,
                      nOuter = 5,
                      varRatio = 0.75,
                      method = "RF",
                      modReturn = TRUE)

actualFit <- MUVR2_actual$ber[model]
actualFit
```

## The null distribution

`H0_test()` takes the fitted model, pulls its data and settings out of it, and
refits it `n` times against a resampled target.

Twenty-five resamplings is far too few for a real analysis — **use at least 100** —
but it keeps this page buildable. Note also that the resampling models inherit the
actual model's settings by default, which is what you want; if computation forces
your hand, you can lower `nRep`, `nOuter` or `varRatio` in the `H0_test()` call.
That broadens the null distribution and inflates p-values, i.e. it errs on the side
of caution.

```{r h0}
nResample <- 25   # use >= 100 for real work

resamplings <- H0_test(MUVR2_actual, n = nResample, type = "resampling")
resampleFit <- resamplings[, model]

summary(resampleFit)
```

## Three ways to get a p-value

The actual fitness has to be compared against that distribution somehow, and there
are three options — none of which is a formality, because they disagree.

**Empirical (`type = "smooth"`)** — the cumulative probability under a smoothed
curve fitted to the resampled distribution. **This is the recommended one.**
Resampled fitness distributions are frequently not Gaussian, and this test does not
assume they are.

```{r perm-smooth, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPerm(actual = actualFit, distribution = resampleFit,
         type = "smooth", curve = TRUE, xlim = c(0, 1), xlab = "BER")
ggplotPerm(actual = actualFit, distribution = resampleFit,
           type = "smooth", curve = TRUE, xlim = c(0, 1), xlab = "BER")
```

**Student's t (`type = "t"`)** — assumes the resampled distribution is
t-distributed and takes the cumulative probability from that. Fine when the
distribution really is Gaussian; misleading when it is not.

```{r perm-t, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPerm(actual = actualFit, distribution = resampleFit,
         type = "t", curve = TRUE, xlim = c(0, 1), xlab = "BER")
ggplotPerm(actual = actualFit, distribution = resampleFit,
           type = "t", curve = TRUE, xlim = c(0, 1), xlab = "BER")
```

**Rank-order (`type = "rank"`)** — the non-parametric test of Szymańska *et al.*
(2012), based purely on where the actual fitness ranks within the null
distribution. Assumption-free, but its resolution is bounded: it can never report a
p-value below 1/`nResample`. With 25 resamplings, that floor is 0.04.

```{r perm-rank, fig.show = "hold", out.width = "49%", fig.width = 5.5}
plotPerm(actual = actualFit, distribution = resampleFit,
         type = "rank", xlim = c(0, 1), xlab = "BER")
ggplotPerm(actual = actualFit, distribution = resampleFit,
           type = "rank", curve = FALSE, xlim = c(0, 1), xlab = "BER")
```

You can put several of them on one plot to see how much they disagree:

```{r perm-multi}
ggplotPerm(actual = actualFit, distribution = resampleFit,
           multiple_p_shown = c("t", "smooth"), xlim = c(0, 1), xlab = "BER")
```

`pPerm()` gives you the p-value without the plot, and `permutationPlot()` /
`ggpermutationPlot()` wrap all of this up if you would rather pass the model and
the `H0_test()` result directly instead of extracting the fitness yourself.

```{r pperm}
pPerm(actual = actualFit, permutation_distribution = resampleFit, type = "smooth")$p
```

## The check most people skip: is the *modelling* overfitting?

Everything above assumes the null distribution is a fair picture of "no
association". But it might not be. The resampled models are still models — they
still fit, and they can still overfit the noise. If the machine learning method
suits the resampled data badly, the whole null distribution shifts, and the p-value
you computed against it is not measuring what you think.

`H0_reference()` gives you the yardstick. It computes the fitness **directly from
the resampled target variables, with no modelling and no prediction at all** — so
by construction it contains no overfitting. It is what the null distribution
*should* look like.

`H0_reference()` draws its histogram and hands back the distribution itself:

```{r h0ref, fig.width = 6, fig.height = 4}
reference <- H0_reference(Y = Yotu, n = 1000, fitness = "BER")

summary(reference)
```

Now compare the two distributions:

- **They line up** → the modelling strategy is not contributing overfitting, and
  your p-value means what it says.
- **They are systematically shifted** → the modelling strategy does not fit this
  data, and the shift is overfitting. Fix that before interpreting any p-value.

```{r compare, fig.width = 7, fig.height = 4}
library(ggplot2)

comparison <- rbind(
  data.frame(BER = as.numeric(resampleFit), source = "Resampled models"),
  data.frame(BER = as.numeric(reference),   source = "Reference (no modelling)")
)

ggplot(comparison, aes(x = BER, fill = source)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = actualFit, linewidth = 0.8) +
  annotate("text", x = actualFit, y = 0, label = "actual model",
           hjust = -0.1, vjust = -0.5) +
  labs(fill = NULL, y = "Density") +
  theme_bw()
```

## Choosing a fitness metric

- **Regression**: Q2.
- **Classification**: BER by default. Number of misclassifications also works, but
  handles imbalanced classes poorly.
- **Multilevel**: any of BER, misclassifications, or Q2.

A closing practical note: start with few resamplings to get a rough sense of where
the actual fitness sits relative to the null, and only then spend the compute on
enough resamplings for a p-value you would put in a paper.
